External priors for the next generation of CMB experiments.

### 2015-09-10 Notes/N_eff_prior_paper.tex
working on the draft:intro

\section{Introduction}\label{sec:intro}
Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe. They will maintain their role also in the near future thanks to the unprecedented level of noise and accuracy of their next generation (S4). 
The recent past is indeed reassuring. 
Every new generation of satellite experiments improved the sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck. With the improved sensitivity we managed to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
The same success characterized ground experiments where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
These series of successes is far from its end. Indeed the next generation of CMB experiments (S4) is now in its planning stage and it will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
As it happened in the past, this new sensitivity will improve our understanding of several areas of astrophysics like neutrinos, inflation, Dark Energy and Dark matter.

%In this work we will mainly focus on two of these: neutrinos and Dark Energy.
The main effects of neutrinos in the CMB can be parametrize by the number of relativistic degree of freedom in the early universe $N_{\rm eff}$ and the total mass of the three neutrino species $M_{\nu}$.
Indeed the number of relativistic species, like neutrinos, in the early Universe alter the expansion history. This can be probed using the CMB by comparing the sound horizon scale, obtained from the CMB peaks positions, with the silk damping scale taking advantage of the fact that they scale differently with the expansion rate $H(z)$ 
%their ratio $r_{d}/r_{s}\propto\sqrt{H}$ 
(see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
This allow CMB experiment to constrain $N_{\rm eff}$. Any deviation from the predicted value $N_{\rm eff}=3.046$ will be a sign of possible extensions of the Standard Model.
The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface. As other low redshift effects however, massive neutrinos modify the CMB through the lensing of the CMB photons by large scale structures. Different neutrino masses result into a different growth of structure which consequently lead to different CMB lensing. The lensing potential can be reconstructed from the CMB itself, opening the possibility of constraining the neutrino mass. 
In the future, according for example to \cite{wu:2014}, CMB alone will constrain $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when baryon acoustic oscillations (BAO) data are also used. The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.

As previously mentioned, CMB is also sensitive to  the Dark energy properties (see \cite{2010MNRAS.405.2639J}). The current generation of experiments measured with percent accuracy the energy density of Dark Energy. The challenge for future experiments is to identify the nature of this mysterious component. The first step will be to measure any deviation in the equation of state from the one predicted by a cosmological constant model. Furthermore CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
Indeed dark energy affect the CMB because due to its effect on the universe's expansion that modifies the distance to the last scattering surface together with a modification  of CMB photons energy through the Integrated Sachs Wolfe Effect. Furthermore different DE models lead to a different growth of large scale structure. However dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. To get competitive constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.

In general, external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters. For this reason the synergy with other experiments is crucial. In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed.
%This can tell us for example which CMB parameter constraints will benefit more from improvement on external priors or if proposed experiments like PIXIE \cite{2011JCAP...07..025K} can indirectly also benefit CMB constraints.

>>>>

\section{Introduction}\label{sec:intro}
Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe. They will maintain their role also in the near future thanks to the unprecedented level of noise and accuracy of their next generation (S4). 
The recent past is indeed reassuring. 
Every new generation of satellite experiments improved the sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck. With the improved sensitivity we managed to extend the scientific outcome of these experiments from a measurement of the universe flatness to a percent level constraint on the parameters of the standard $\Lambda$CDM model.
The same success characterized ground experiments where the progression from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
These series of successes is far from its end. Indeed the next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
As it happened in the past, this new sensitivity will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.

The cosmological dependence to neutrinos is primarily due to their contribution to the number of relativistic species $N_{\rm eff}$ in the early phase of the universe  together with the role they play, given their non zero mass $M_{\nu}$, into the late growth of cosmic structures.
Because relativistic species, like neutrinos, present in the early Universe are the main drivers of the cosmic expansion their number is encapsulated in the expansion rate $H(z)$. This rate can be powerfully tested using the CMB. Mainly by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale and will unequivocally test with data the predicted value of the standard model $N_{\rm eff}=3.046$.
The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface. As other low redshift effects however, massive neutrinos modify the CMB by altering the growth of the large scale structure responsible for the the lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing. Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise drastically improving the lensing reconstruction. This will turn in a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses. 
To put this progress in prospective, CMB alone will constrain the relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when we introduce priors from baryon acoustic oscillations (BAO) data \cite{wu:2014}. The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.

As previously mentioned, CMB is also sensitive to the properties of Dark energy (see \cite{2010MNRAS.405.2639J}). The current generation of experiments measured with percent accuracy the energy density of Dark Energy. The challenge for future experiments is to reveal the nature of this mysterious component. For example a crucial step will be to investigate any possible deviations in the equation of state from the value $w=-1$ predicted by a cosmological constant model. Furthermore CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
Indeed dark energy affects the CMB because it alters the universe's expansion consequently changing the distance to the last scattering surface. Furthermore different DE models lead to a different growth of large scale structure that can be inferred from the CMB lensing. However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. Similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.

In general, external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters. For this reason the synergy with other experiments is crucial. In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study in detail the dependance of CMB future constrains on the external priors assumed.
